---
title: "Taxonomy of statistics"
author: "NRES 710"
date: "Fall 2022"
output: 
  html_document: 
    theme: yeti
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, cache = TRUE)

```

```{r echo=FALSE}

############################################################
####                                                    ####  
####  NRES 710, Lecture 3                               ####
####                                                    ####
####  Kevin Shoemaker and Ben Sullivan                  #### 
####  University of Nevada, Reno                        ####
####                                                    #### 
############################################################


############################################################
####  Chi-squared tests etc.                            ####
############################################################


```



## Overview of statistical methods

Okay before we delve further into t-tests, let's first give a broad overview of the analyses we will cover in this class (at least a tiny bit) and in which cases they are most appropriate:

### Continuous response variable

If your response variable is continuous (ratio, interval), common classical parametric statistical analyses include: t-tests, ANOVA, linear regression. Each of these tests is associated with non-parametric alternatives. In each case, we are interested in testing if the *population mean* of the continuous response variable is affected by the predictor variables (null hypothesis: nope, there is no effect!). Predictor variables, in turn, can be continuous or categorical (factor variables in R).   

#### Continuous response, binary categorical predictor

If your categorical predictor is binary (two levels), you can use a **two-sample t-test**. The non-parametric alternative is the **Mann-Whitney test**.

Example: Are females larger than males? (null hypothesis: no difference)

If your categorical predictor has more than two levels, you can use an **ANalysis Of Variance (ANOVA)** followed by *pairwise comparisons* to test which categories differ from one another. You can visualize these relationships with (e.g.) a boxplot. The non-parametric alternative is the **Kruskal-Wallis test**.  

**Q**: What is the predictor variable for a one-sample t-test?

#### Continuous response, continuous predictor

If your predictor variable is continuous, you can use **linear regression**. If you just want to know if two variables are correlated but you are not interested in modeling one (the response) as a function of another (the predictor) you can run a **Pearson correlation test**. The non-parametric alternative is a **Spearman correlation test**.    

Example: What is the relationship between tree diameter and age- can tree diameter be used to effectively predict the age of a tree?  

#### Continuous response, both continuous and categorical predictors

In this case you can use **Analysis of Covariance (ANCOVA)** or just use **multiple linear regression**. In truth, linear regression and ANOVA/ANCOVA are two sides of the same coin. You can run both analyses using the workhorse of linear modeling in R, the 'lm' function. Basically, if you have a continuous response variable, you can use the 'lm' function! The non-parametric alternative might be something like a **regression tree** analysis. 

### Discrete (count) response variable

With a discrete count response, you can use the same techniques as with a continuous response OR you can use **generalized linear models** with a Poisson error distribution (or other discrete probability distribution).

### Categorical response variable

If your response variable is categorical (factor variable, ordinal variable, binary variable) then your choice of statistical methods changes:

#### Categorical response variable, categorical predictor

If both your response variable and your predictor variable are categorical, then you can use a **chi-squared test** or a **Fisher exact test** to test for an association between the two variables. You may first summarize your data as a *contingency table* like we did with the 'lady tasting tea' example.

Example: test for an association between salamander color morph (melanistic vs wild-type) and mating behavior ('sneaker' vs territory holder).  
 
#### Categorical response variable, continuous predictor 

If your response variable is binary (true/false, two levels) and your predictor variable is continuous, you can use **logistic regression** (which is a type of *generalized linear model (GLM)*). 

If your categorical response variable has more than two levels, you can use **multinomial logistic regression** (for categorical responses) or **ordinal logistic regression** (for ordinal responses). 
 

[--go to next lecture--](LECTURE4.html)













