---
title: "Next Steps"
author: "NRES 710"
date: "Fall 2020"
output: 
  html_document: 
    theme: yeti
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE, cache = TRUE)

```

```{r echo=FALSE}

############################################################
####                                                    ####  
####  NRES 710, Lecture 6                               ####
####                                                    ####
####  Kevin Shoemaker and Ben Sullivan                  #### 
####  University of Nevada, Reno                        ####
####                                                    #### 
############################################################


############################################################
####  Next Steps                                        ####
############################################################


```

## Download the R code for this lecture!

To follow along with the R-based lessons and demos, [right (or command) click on this link and save the script to your working directory](LECTURE7.R)

## Overview: Generalized Linear Models

Generalized Linear Models (GLM) are not generally covered in intro stats classes, but they are so common in ecology and environmental science that you really need to at least know that these models exist!



--End of lectures--

















